Specialized Cost Function for Maximizing Body Area Networks Lifetime Using Global Routing Algorithms

### 2011-09-13 docs/thesis/paper.tex
Updated paper.tex and added thesis.tex

\documentclass{article}
\begin{document}
\title{Specialized Cost Function for Maximizing Body Area Networks Lifetime Using Global Routing Algorithms}
\author{Alvaro Prieto}
\maketitle

@@

There are several routing protocols designed specifically for WBANs. Braem et al. developed the CICADA protocol, which consists of a spanning tree architecture with a time-division scheme for transmission scheduling~\cite{protocol:CICADA}. The primary downside to this method is that nodes closer to the root will deplete their energy source faster due to the need to relay messages from children nodes. Other protocols focus on the reliability of the network$\ll$citation needed$\gg$. Dynamic networks are a problem with WBANs. If sensors are placed in the extremities, the network will change as the body changes positions. Quwaider et al. developed a protocol tolerant to network changes~\cite{routing:storeandforward}. They propose a store-and-forward method that maximizes the likelihood of a packet reaching its destination. Each packet is stored by multiple devices and retransmitted, which consumes more power. One solution for this problem was proposed by Ehyaie et al.~\cite{relay:networklife}, it consists of using dedicated, non-sensing, relay nodes with larger power sources. While this method increases the network lifetime$\ll$define term$\gg$, it requires more hardware. Finally, Nabi et al. propose a similar store-and-forward method, but integrates Transmit Power Adaptation (TPA)~\cite{relay:transmitpoweradaptation}. Nodes keep track of neighbors and use power control to use the smallest transmit power while maintaining a specific link quality.

@@

@@

Each device's energy use is accumulated as shown in Equation~\ref{eqn:accumulated_energy}, where $j$ denotes the device ID, $i$ is the current polling round, $\alpha$ is the channel attenuation, and $RSSI^T$ is the target RSSI. The current energy used is incremented by the energy used to transmit a single packet while maintaining $RSSI^T$. The channel attenuation for the selected link,$\alpha$, can be seen in Equation~\ref{eqn:alpha}, where RSSI is the received power and $P_tx$ is the transmitted power. The actual link cost is computed by calculating the energy used by a device \emph{if} that link is selected and subtracting the average energy used across all devices. Equation~\ref{eqn:link_cost} is normalized by the current polling round $i$.
\begin{equation}
\label{eqn:alpha}
\alpha = \frac{RSSI}{P_{tx}}
\end{equation}
\begin{equation}
\label{eqn:accumulated_energy}
E^j_{acc_{(i)}} = E^j_{acc_{(i-1)}} + \frac{RSSI^T}{\alpha}
\end{equation}
\begin{equation}
\label{eqn:link_cost}
C^j_{l,k} = \left[ \left(\frac{ E^l_{acc_{(i-1)}} + \frac{RSSI^T}{\alpha_{l,k}}}{i}\right) - \frac{\sum_jE^l_{acc_{(i-1)}}}{i-1} \right]
\end{equation}

@@

\section{Hardware Implementation}
The hardware platform used for this test consists of the Texas Instruments (TI) EZ430-RF2500. This device includes both an MSP430F2274 microcontroller along with a CC2500 2.4GHz transceiver. A single device, labeled the Access Point (AP), is connected via a USB to Serial link to the host computer. All other devices, labeled End Devices (EDs), are battery powered. In this implementation, the AP acts only as a bridge between the host and the end devices. All routing and power control calculations are done on the host.

A normal cycle of operations with $n$ devices works as follows:
\begin{enumerate}
\item AP sends synchronization beacon which includes routing and power control tables.
\item Each ED transmits its own RSSI table back to the AP, while simultaneously listening to other ED messages and storing the received power.
\item Once all EDs have transmitted their data, the AP sends a table with the RSSI data from all devices.
\item The host uses the RSSI table to compute the routes along with the required powers to meet the selected links.
\item The host sends both routing and power tables back to the AP so that a new cycle may begin.
\end{enumerate}

To minimize the number of control packets from being transmitted, the EDs are not individually polled. The only control packet that is sent is the synchronization beacon, which also carries the routing and power tables. Once the EDs are synchronized, they transmit on a pre-defined schedule to avoid collisions. Each ED has a network ID ranging from 1 to $n$. The time between synchronization packets is divided into time-slots than are used by each ED. The time slot used depends on the network ID of each device. This removes the need to come up with time schedules during runtime.

@@

>>>>

\documentclass{article}
\usepackage{graphicx}
\begin{document}
\title{Specialized Cost Function for Maximizing Body Area Networks Lifetime Using Global Routing Algorithms}
\author{Alvaro Prieto}
\maketitle

@@

There are several routing protocols designed specifically for WBANs. Braem et al. developed the CICADA protocol, which consists of a spanning tree architecture with a time-division scheme for transmission scheduling~\cite{protocol:CICADA}. The primary downside to this method is that nodes closer to the root will deplete their energy source faster due to the need to relay messages from children nodes. Other protocols focus on the reliability of the network~\cite{routing:storeandforward}. Dynamic networks are a problem with WBANs. If sensors are placed in the extremities, the network will change as the body changes positions. Quwaider et al. developed a protocol tolerant to network changes~\cite{routing:storeandforward}. They propose a store-and-forward method that maximizes the likelihood of a packet reaching its destination. Each packet is stored by multiple devices and retransmitted, which consumes more power. One solution for this problem was proposed by Ehyaie et al.~\cite{relay:networklife}, it consists of using dedicated, non-sensing, relay nodes with larger power sources. While this method increases the network lifetime$\ll$define term$\gg$, it requires more hardware. Finally, Nabi et al. propose a similar store-and-forward method, but integrates Transmit Power Adaptation (TPA)~\cite{relay:transmitpoweradaptation}. Nodes keep track of neighbors and use power control to use the smallest transmit power while maintaining a specific link quality.

@@

\begin{figure}[htb]
\begin{center}
\includegraphics[width=0.5\textwidth]{figures/body.pdf}
\end{center}
\caption{Experimental Setup}
\label{fig:body}
\end{figure}

@@

Each device's energy use is accumulated as shown in Equation~\ref{eqn:accumulated_energy}, where $j$ denotes the device ID, $i$ is the current polling round, $\alpha$ is the channel attenuation, and $RSSI^T$ is the target RSSI. The current energy used is incremented by the energy used to transmit a single packet while maintaining $RSSI^T$. The channel attenuation for the selected link,$\alpha$, can be seen in Equation~\ref{eqn:alpha}, where RSSI is the received power and $P_tx$ is the transmitted power. The actual link cost is computed by calculating the energy used by a device \emph{if} that link is selected and subtracting the average energy used across all devices. Equation~\ref{eqn:link_cost} is normalized by the current polling round $i$.
\begin{equation}
\label{eqn:alpha}
\alpha = \frac{RSSI}{P_{tx}}
\end{equation}
\begin{equation}
\label{eqn:accumulated_energy}
E^j_{(i)} = E^j_{(i-1)} + \frac{RSSI^T}{\alpha}
\end{equation}
\begin{equation}
\label{eqn:link_cost}
C^j_{l,k} = \left[ \left(\frac{ E^l_{(i-1)} + \frac{RSSI^T}{\alpha_{l,k}}}{i}\right) - \frac{\sum_jE^l_{(i-1)}}{i-1} \right]
\end{equation}

\begin{equation}
\label{eqn:link_cost}
C^j_{l,k} = \frac{RSSI^T}{\alpha_{l,k}} \times \left( \frac{1+\left(\frac{E^j_i}{E^{min}_{i}} \right)^c}{2} \right)
\end{equation}

@@

\section{Hardware Implementation}
\subsection{Hardware Platform}
The hardware platform used for this test consists of the Texas Instruments (TI) EZ430-RF2500. This device includes both an MSP430F2274 microcontroller along with a CC2500 2.4GHz transceiver. A single device, labeled the Access Point (AP), is connected via a USB to Serial link to the host computer. All other devices, labeled End Devices (EDs), are battery powered. In this implementation, the AP acts only as a bridge between the host and the end devices. All routing and power control calculations are done on the host.

The host is a laptop computer with an Intel(R) Core(TM) i5-2410M CPU and 4GB of RAM. Eight EDs were placed on a 170cm, 70kg male subject as shown in Figure~\ref{fig:body}. The host was carried on a backpack and connected to the AP via a USB cable. The data was sampled, and the routing algorithm was run, at a rate of 5Hz. This sampling rate is appropriate as long as the subject, or its environment, does not move faster than this.$\ll$What was the name for this?$\gg$ If a data needs to be sampled at a faster rate, some hardware and firmware changes might be necessary, but the host software will still work.

\subsection{Cycle of Operations}
A normal cycle of operations with $n$ devices works as follows:
\begin{enumerate}
\item AP sends synchronization beacon which includes routing and power control tables.
\item Each ED transmits its own RSSI table back to the AP, while simultaneously listening to other ED messages and storing the received power.
\item Once all EDs have transmitted their data, the AP sends a table with the RSSI data from all devices.
\item The host uses the RSSI table to compute the routes along with the required powers to meet the selected links.
\item The host sends both routing and power tables back to the AP so that a new cycle may begin.
\end{enumerate}

\subsection{Details}
To minimize the number of control packets from being transmitted, the EDs are not individually polled. The only control packet that is sent is the synchronization beacon, which also carries the routing and power tables. Once the EDs are synchronized, they transmit their data on a pre-defined schedule to avoid collisions. Each ED has a network ID ranging from 1 to $n$. The time between synchronization packets is divided into time-slots than are used by each ED. The time slot used depends on the network ID of each device. This removes the need to come up with time schedules during runtime.

@@

\begin{figure}[htb]
\begin{center}
\includegraphics[width=0.5\textwidth]{figures/walk2-c100.pdf}
\end{center}
\caption{Walking (C=100)}
\label{fig:walk2-c100}
\end{figure}

\begin{figure}[htb]
\begin{center}
\includegraphics[width=0.5\textwidth]{figures/walk2-c0.pdf}
\end{center}
\caption{Walking (C=0)}
\label{fig:walk2-c0}
\end{figure}
